Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-09-08 revised_main.tex
Updates from Overleaf

We provide a review of past works in order to understand what is lacking, a nomenclature for different subtypes of sound-matching, and illuminate major perennial issues. In particular, we are concerned with the issues of \LossSelect~and \SynthSelect. The former refers to the search for an optimal loss function, lack of novel algorithms, and the---perhaps needless---focus on outperforming state of the art (\gls{SOTA}), while the latter refers to the lack of diversity in synthesis methods. As a consequence of these problems, we find that not much is known about the interaction between different methods of synthesis and different loss functions. How much performance variation can we expect?

The main hypothesis here is that \textit{the performance of a similarity measure (or loss function) is influenced by other factors in the environment, particularly the method of synthesis}. While there have been many works and experiments comparing the accuracy of loss functions~\cite{vahidi2023mesostructures,turian2020sorry,engel2020ddsp,uzrad2024diffmoog,han2023perceptual,masuda2021soundmatch,turian2020sorry,bruford2024synthesizer}, we note that claims regarding the effectiveness of one function versus another have consistently been made in limited contexts that may not generalize to other settings. Beyond this central hypothesis, we also investigate three additional research questions: (Q1) To what extent do automatic evaluation metrics agree with manual listening tests? (Q2) Can novel loss functions utilizing Dynamic Time Warping (DTW) and Scale-Invariant Mean Squared Error (SIMSE) provide advantages over standard loss functions? (Q3) Is an iterative differentiable optimization a viable strategy for design of sound-matching experiments?

@@

 The iterative approach (commonly used in earlier works) better mimics the manual process of sound design (recursively listening to synthesizer outputs and adjusting the parameters accordingly) while a differentiable environment allows access to the loss function gradients that can be used to better understand the nature of the problem. We define four differentiable synthesizers (each showcasing a fundamental method of synthesis used in modern synthesizers) and pair them with four different loss functions (two established methods, one utilizing DTW, and one utilizing SIMSE). We evaluate the final similarity with two different automatic methods, as well as manual listening tests conducted by two of the authors. We compare the distributions of the evaluation scores in order to rank the loss functions from best to worst.

The results of these experiments show that when accounting for different synthesizers, there is no consistent choice of sound similarity measure for optimal performance. Regarding Q1, We see a large overlap between automatic and manual evaluation methods, but continue to encourage manual listening tests as the gold standard for audio evaluation as extreme disagreements can occur. In the case of Q2, we confirm that \DTWEnv outperforms other methods when envelope modulation plays a central role, and \SIMSESpec is the best performer in simple subtractive synthesis. For Q3, we found the declaration of differentiable sound-matching experiments using Jax and transpiled Faust code to be a convenient and resource efficient, and and overall viable general strategy for sound-matching in differentiable settings. These results imply that rather than aiming to outperform the SOTA and finding a universal best performer, more effort could be placed towards diversifying the existing similarity measures of sound as well as conducting experiments using a variety of synthesis methods. Sound design is an inherently creative endeavor, and would likely benefit from a diversified range of solutions for different approaches.

>>>>

We provide a review of past works in order to understand what is lacking, a nomenclature for different subtypes of sound-matching, and illuminate major perennial issues. In particular, we are concerned with the issues of \LossSelect~and \SynthSelect. The former refers to the search for an optimal loss function, lack of novel algorithms, and the---perhaps needless---focus on outperforming state of the art (\gls{SOTA}), while the latter refers to the lack of diversity in synthesis methods. As a consequence of these problems, we find that not much is known about the interaction between different methods of synthesis and different loss functions. Is there a universally best performing loss function for audio?

The main hypothesis here is that \textit{the performance of a similarity measure (or loss function) is influenced by other factors in the environment, particularly the method of synthesis}. While there have been many works and experiments comparing the accuracy of loss functions~\cite{vahidi2023mesostructures,turian2020sorry,engel2020ddsp,uzrad2024diffmoog,han2023perceptual,masuda2021soundmatch,turian2020sorry,bruford2024synthesizer}, we note that claims regarding the effectiveness of one function versus another have consistently been made in limited contexts that may not generalize to other settings. Beyond this central hypothesis, we also investigate three additional research questions: (Q1) To what extent do automatic evaluation metrics agree with manual listening tests? (Q2) Can novel loss functions utilizing Dynamic Time Warping (DTW) and Scale-Invariant Mean Squared Error (SIMSE) provide advantages over standard loss functions, and if so, when? (Q3) Is an iterative differentiable optimization a viable strategy for design of sound-matching experiments?

@@

 We take an \textit{iterative} and \textit{differentiable} approach to defining our sound-matching experiments. The iterative approach (commonly used in earlier works) better mimics the manual process of sound design (recursively listening to synthesizer outputs and adjusting the parameters accordingly) while a differentiable environment allows access to the loss function gradients that can be used to better understand the nature of the problem. We define four differentiable synthesizers (each showcasing a fundamental method of synthesis used in modern synthesizers) and pair them with four different loss functions (two established methods, one utilizing DTW, and one utilizing SIMSE). We evaluate the final similarity with two different automatic methods, as well as manual listening tests conducted by two of the authors. Using the Non-parametric Scott-Knott tests~\cite{tantithamthavorn2017mvt,tantithamthavorn2018optimization}, compare the distributions of the evaluation scores in order to rank the loss functions from best to worst.

The results of these experiments show that when accounting for different synthesizers, there is no consistent choice of sound similarity measure for optimal performance. Regarding Q1, We see a large overlap between automatic and manual evaluation methods, but continue to encourage manual listening tests as the gold standard for audio evaluation as extreme disagreements can occur. In the case of Q2, we confirm that \DTWEnv~outperforms other methods when envelope modulation plays a central role, and \SIMSESpec~is the best performer in simple subtractive synthesis. For Q3, we found the declaration of differentiable sound-matching experiments using Jax and transpiled Faust code to be a convenient and resource efficient, and and overall viable general strategy for sound-matching in differentiable settings. These results imply that rather than aiming to outperform the SOTA and finding a universal best performer, more effort could be placed towards diversifying the existing similarity measures of sound as well as conducting experiments using a variety of synthesis methods. Sound design is an inherently creative endeavor, and would likely benefit from a diversified range of solutions for different approaches.
